---
title: "session04 - homework - solutions"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, ggeffects, patchwork)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction to data

We will be working with perception data from one of my own experiments [@roettger2014assessing].

In this study, we investigated incomplete neutralization (IN). IN refers to phenomena in which a phonological neutralization is phonetically incomplete. In this paper, researchers looked at final devoicing in German. In German, the word "Räder" (*engl.* 'wheels') has a voiced /d/, but when forming the singular "Rad" (*engl.* 'wheel'), the /d/ is in word-final position and is devoiced, i.e. loses its phonological voicing and thus is supposed to be identical to the /t/ sound in "Rat" (*engl.* 'council'). However, when measuring the /d/ in "Rad" and the /t/ in "Rat", they are phonetically not identical. There are small but consistent differences between them. In experiment 4 of the paper, the researchers test whether these small acoustic differences can be perceived by naive listeners.

They showed 16 listeners 24 different pseudoword pairs produced by 16 different speakers and measured both reaction time as well as accuracy of correctly identifying the speaker-intended voicing category.

Let's load in the data and have a look:

```{r}
#| label: load-in

incomplete <- read_csv("data/incomplete_perception.csv") |> 
  # preprocessing the data such that everything that ought to be a factor is one
  mutate(listener = as.factor(listener),
         voice = as.factor(voice), 
         item_pair = as.factor(item_pair),
         correct_voicing = as.factor(correct_voicing),
         chosen_voicing = as.factor(chosen_voicing)
         )

# have a look
incomplete

#i do not know why i cannot for the life of me make the file open so i just had to import it manually :((

incomplete_perception

incomplete <- incomplete_perception |> 
  # preprocessing the data such that everything that ought to be a factor is one
  mutate(listener = as.factor(listener),
         voice = as.factor(voice), 
         item_pair = as.factor(item_pair),
         correct_voicing = as.factor(correct_voicing),
         chosen_voicing = as.factor(chosen_voicing)
         )

incomplete
```

These are the variables that you will find within the data:

-   `order`: trial order for each listener (1-392)
-   `listener`: unique identifier for each listener (n = 16)
-   `age`: listeners' age
-   `item_pair`: unique identifier for each word pair (e.g. pruk vs. prug, n = 24)
-   `voice`: unique identifier for each speaker voice (n = 16)
-   `RT` = reaction time in ms.
-   `ACC` = accuracy (binary, 1 = correct, 0 = incorrect)
-   `correct_voicing` = whether the audio stimuli contained an underlyingly voiced or voiceless stop in word-final position
-   `chosen_voicing` = the voicing category that the listeners selected in a 2-alternative forced-choice.

### (A) Run a mixed effects model

In their paper, they described different ways how to analyse the perception data. Among other things, they use a generalized linear mixed effects model. While not clearly articulated in the paper, they were mainly interested in whether listeners chose the intended voicing category more often than the unintended one. One way to test this hypothesis is to predict the `chosen_voicing` category by the `correct_voicing` category (i.e. the one that was intended by the speaker).

Build that model with the appropriate random effect structure, taking into account all important grouping variables.

(NOTE: if you encounter the following error message, ignore for now and rerun the model until it does not give this error: *Error in pwrssUpdate(pp, resp, tol = tolPwrss, GQmat = GQmat, compDev = compDev, : Downdated VtV is not positive definite*)

```{r}
#| label: glmer_incomplete

# YOUR CODE HERE

glmer_incomplete <- glmer(chosen_voicing~correct_voicing + (1|listener) #most obviously different intercepts for different listeners
                          + (1|item_pair) + (1|voice), #speaker voice may affect how different phonemes sound, and different phoneme pairs may have different effects
                          data=incomplete, family=binomial)
#i feel like i should add random slopes for listeners somehow but i am a bit unsure how to go about it... item_pair? correctness? voice? all?? TT

glmer_incomplete

```

### (B) Interpret the model output

Interpret the model coefficients and calculate p-values for the relevant predictor to test the hypothesis by comparing the full model (with the predictor) with the corresponding null model (without the predictor, i.e. just substitute the fixed effect predictor with 1).

```{r}
#| label: interpretation

# YOUR CODE HERE

summary(glmer_incomplete)

#interpreting random effects: voice (speaker producing the items) accounts for the most variance in data. this indicates that whether a phoneme was categorised correctly or not varied between speakers, and that some speakers may have had clearer distinctions between voice and voiceless phonemes than others. item_pair explains a bit less variance, but also indicates that the specific phoneme pairs differed in correctness, which might mean that some phoneme pairs were easier/harder to distinguish. lastly, listener did not explain a lot of variance. this tells us that any differences in correctness were not caused by there being a difference in the skill of the listeners to identify voiced/voiceless phonemes, and that the different listeners performed similarly overall.
#interpreting fixed effects: the intercept refers to the log odds of having chosen correctly (voiced) when the item was voiced, which is -0.03. I honestly do not know what that means but it does not look good. correct_voicing voiceless refers to the log odds of having chosen "voiced" when the item was voiceless, which is 0.52. Idk what that means either (yet).

source("https://sebastiansauer.github.io/Rcode/logit2prob.R") #thank you github

logit2prob(-0.03262)

#output: 0.49. interpretation: for voiced items, listeners chose correctly 49% of the time. This likely means they were guessing, since they only had 2 answers to choose from, and a random guess in this situation would give us the correct result 50% of the time.

-0.03262+0.52195 #getting the log odds for voiceless, as the former number is the log odds for voiceless as it differs from the log odds for voiced

logit2prob(0.48933)

#output: 0.62. interpretation: for voiceless items, listeners chose the "voiced" category 62% of the time, aka choosing incorrectly. This means they only correctly identified the voiceless phoneme 38% of the time.
#The listeners were worse at identifying voiceless phonemes than voiced ones, which is interesting given that they performed at chance for the voiced ones. This means they could have performed better at identifying the voiceless phonemes if they had been randomly guessing.

full_model <- glmer_incomplete
null_model <- lme4::glmer(chosen_voicing ~ 1 + (1 | listener) + (1 | item_pair) + (1 | voice), data = incomplete, family = binomial)

?anova

anova(null_model, full_model)
#is there no way to make this digestable at all
#interpretation: full_model is statistically sagnificant, p-value smaller than 0.05


```

### (C) Plot predictions

Plot the model predictions alongside a measure of uncertainty (standard error or 95% CI).

Estimating variance components for linear mixed effects models is mathematically not as straightforward as it is for simple linear models. We shortcut manual extraction for now and use the beautiful `ggeffects` package. Check their `predict_response()` function and use it to derive predictions and plot them.

```{r}
#| label: plot_prediction_in

# YOUR CODE HERE

predict <- ggeffects::predict_response(full_model, terms = "correct_voicing")

#i do not know what that warning means

predict2 <- ggeffects::predict_response(full_model, terms = "correct_voicing", bias_correction = TRUE)

print(predict)
print(predict2)
#i am not entirely sure what the bias correction did but given that "predict" has my numbers (49, 62) while predict2 has something else (50, 53), i will go with predict

plot(predict)

library(ggplot2)

plot_prediction_in <- plot(predict) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.4) +
  labs(title = "Model Predictions of 'Voiced' with 95% CI",
       x = "Correct Voicing Category",
       y = "Predicted Probability") +
  theme_minimal()

plot_prediction_in

```

### (D) OPTIONAL Plot variation across grouping variables

Compare the descriptive averages for the critical relationship for your grouping variables and describe what's going on.

```{r}
#| label: plot_descriptive_in

# YOUR CODE HERE

```

We can see massive variation, particularly for different item pairs and different speaker voices.

### (E) Use Github to upload your analysis

-   fork https://github.com/timo-b-roettger/ling4136-advanced-stats
-   clone your forked repo onto your computer
-   store your solutions in this script as `YOURLASTNAME_04_homework.qmd`
-   move your script into `/homework/04_homework/`
-   commit and push changes to your forked repo
-   add me as a collaborator to your repository and send the link to your homework through Canvas

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
